---
title: "getCordex - Get projection data from CORDEX"
author: Sylvain Schmitt
date: April 18, 2023
output:
  github_document: 
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(tidyverse)
theme_set(bayesplot::theme_default())
opts_chunk$set(echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
               cache = F, cache.lazy = F)
```

[`singularity` & `snakemake`](https://github.com/sylvainschmitt/snakemake_singularity) workflow to download projection data from CORDEX.

```{r dag, fig.cap="Workflow."}
knitr::include_graphics("dag/dag.svg")
```

# Data

**[CORDEX](https://cordex.org/): Cordinated Regional Climae Downscaling Experiment**

*The CORDEX vision is to advance and coordinate the science and application of regional climate downscaling through global partnerships.*

Data are focused here on South-America (**SAM**), but WGET scripts can be added in the `wget` folder.

We particularly focused on domain **SAM-22** with following variables:
**Near-Surface Air Temperature** (tas)
**Near-Surface Relative Humidity** (hurs),
**Near-Surface Wind Speed** (sfcWind),
**Precipitation** (pr),
**Surface Downwelling Shortwave Radiation** (rsds), &
**Surface Upwelling Shortwave Radiation** (rsus).
We restricted outputs at a **3hr** time-frequency and used the last available downscaling (**v1**).
We used 3 types of experiments: one historical (**historical**) and 
two Representative Concentration Pathway RCP 2.6 & 8.5 (**rcp26** & **rcp85**).
We had 5 driving models available 
(**ECMWF-ERAINT**, **MOHC-HadGEM2-ES**, **MPI-M-MPI-ESM-LR**, **MPI-M-MPI-ESM-MR**, **NCC-NorESM1-M**) 
coupled with 2 RCM model (**REMO2015** & **RegCM4-7**).

Data are retrieve from the [Institut Pierre-Simon Laplace node](https://esgf-node.ipsl.upmc.fr/search/cordex-ipsl/).
You need first to [create an account](https://esgf.github.io/esgf-user-support/user_guide.html#create-an-account) 
on this page ([create account](https://esgf-node.ipsl.upmc.fr/user/add/?next=http://esgf-node.ipsl.upmc.fr/search/cordex-ipsl/) link at corner right).

Then you'll need to register credentials locally to use the workflow.
A help page is available here: https://esgf.github.io/esgf-user-support/user_guide.html?highlight=credentials%20pem#access-data-with-the-command-line-via-opendap.
In linux, you need `myproxy-logon` installed and to run this command line (with your user name):

```{bash, eval=FALSE, echo=TRUE}
myproxy-logon -sesgf-node.ipsl.upmc.fr -l {user_name} -b -T -t 72 -o ~/.esg/credentials.pem
```

To run the workflow on a cluster, you can simply copy your local credentials to the cluster.
For instance:
```{bash, eval=FALSE, echo=TRUE}
cp ~/.esg/credentials.pem /my_cluster/
```

# Installation

- [x] Python ≥3.5
- [x] Snakemake ≥5.24.1
- [x] Golang ≥1.15.2
- [x] Singularity ≥3.7.3
- [x] This workflow

```{bash, eval=F, echo=T}
# Python
sudo apt-get install python3.5
# Snakemake
sudo apt install snakemake`
# Golang
export VERSION=1.15.8 OS=linux ARCH=amd64  # change this as you need
wget -O /tmp/go${VERSION}.${OS}-${ARCH}.tar.gz https://dl.google.com/go/go${VERSION}.${OS}-${ARCH}.tar.gz && \
sudo tar -C /usr/local -xzf /tmp/go${VERSION}.${OS}-${ARCH}.tar.gz
echo 'export GOPATH=${HOME}/go' >> ~/.bashrc && \
echo 'export PATH=/usr/local/go/bin:${PATH}:${GOPATH}/bin' >> ~/.bashrc && \
source ~/.bashrc
# Singularity
mkdir -p ${GOPATH}/src/github.com/sylabs && \
  cd ${GOPATH}/src/github.com/sylabs && \
  git clone https://github.com/sylabs/singularity.git && \
  cd singularity
git checkout v3.7.3
cd ${GOPATH}/src/github.com/sylabs/singularity && \
  ./mconfig && \
  cd ./builddir && \
  make && \
  sudo make install
# detect Mutations
git clone git@github.com:sylvainschmitt/getCordex.git
cd getCordex
```

# Usage

## Locally

```{bash, eval=F, echo=T}
snakemake -np -j 1 --resources mem_mb=10000 # dry run
snakemake --dag | dot -Tsvg > dag/dag.svg # dag
snakemake -j 1 --resources mem_mb=10000 # run
```

## HPC

### Muse

```{bash, eval=F, echo=T}
module load snakemake # for test on node
snakemake -np # dry run
sbatch job_muse.sh # run
```

### Genologin

```{bash, eval=F, echo=T}
module load bioinfo/snakemake-5.25.0 # for test on node
snakemake -np # dry run
sbatch job_genologin.sh # run
```

# Workflow

### [download](https://github.com/sylvainschmitt/getCordex/blob/main/rules/download.smk)

Download data from the [Institut Pierre-Simon Laplace node](https://esgf-node.ipsl.upmc.fr/search/cordex-ipsl/)
using wget scripts from the [`wget/` folder](https://github.com/sylvainschmitt/getCordex/tree/main/wget).

### [extract_var](https://github.com/sylvainschmitt/getCordex/blob/main/rules/extract_var.smk)

* Script: [`extract_var.R`](https://github.com/sylvainschmitt/getCordex/blob/main/scripts/extract_var.R)

Extract variables from the netcdf files in parallel into one tsv table per experiment and variable.

### [join_exp](https://github.com/sylvainschmitt/getCordex/blob/main/rules/join_exp.smk)

* Script: [`join_exp.R`](https://github.com/sylvainschmitt/getCordex/blob/main/scripts/join_exp.R)

Join variables from the same experiment into one single tsv table.

### [reformat](https://github.com/sylvainschmitt/getCordex/blob/main/rules/reformat.smk)

* Script: [`reformat.R`](https://github.com/sylvainschmitt/getCordex/blob/main/scripts/reformat.R)

Reformat experiment variables into variables used in the TROLL model:

* rainfall (rainfall in mm) = 60 * 60 * time_freq_hr * pr (Precipitation in kg/m/s)
* snet (absorbed short-wave radiation in w/m2) = rsds (Surface Downwelling Shortwave Radiation in W/m2) - rsus (Surface Upwelling Shortwave Radiation in W/m2)
* temperature (temperature in degrees C) = tas (Near-Surface Air Temperature in Kelvin) - 273.15
* vpd (vapour pressure deficit in Pa) = rh_to_vpd[ hurs(Near-Surface Relative Humidity in %), temperature ]
* ws (wind speed in m/s) = sfcWind (Near-Surface Wind Speed in m/s)

### [plot](https://github.com/sylvainschmitt/getCordex/blob/main/rules/plot.smk)

* Script: [`plot.R`](https://github.com/sylvainschmitt/getCordex/blob/main/scripts/plot.R)

Plot TROLL variables into hourly, daily, monthly and yearly variation for each variable and experiment.
